GMSE: an R package for generalised management strategy evaluation

### 2018-04-18 notebook/ms/SI3.Rmd
Edits to SI

@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016]. They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`. Similarly, @Johnson2018 estimated intrinsic reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`. The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`). And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`).

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess confict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer simulate by looping time steps through `gmse_apply`, which allows more flexibility for incorporating custom submodels and dynamically adjusting parameter values. Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above. We also set `manager_budget = 100000` and `user_budget = 100000`. Further, we consider the case in of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be intpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`). Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the advanced GMSE options using `gmse_apply`). Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`. We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`). We further assume that the population is well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe two landscape cells in every direction (`agent_view = 2`). All other parameter values are set to GMSE defaults.

@@

Below, we first allow only the culling of geese by users and plot the dynamics of the social-ecological system from a single simulation. Next, we run the same simulation but allow users to scare geese from their landscape cells; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system. We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.

@@

Figure 1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy. In the case of the simulation above, managers quickly set a policy of high cost for culling, which leads to a rise in the goose population and a decrease in crop yield for farmers. After roughly 20 time steps, the goose population rises above the manager target, at which point the manager becomes more permissive of culling and the cost of culling for users therefore declines. In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese. We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function in GMSE.

@@

Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action. When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production. These time periods represent instances in which the manager is not at very permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway. In the case of the above simulation of potential conservation conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target. Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance. It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population is achieved (Figure 1). Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.

We can model the consequences for goose population dynamics, agricultural produciton, and conservation conflict when scaring is a policy option available to the manager. The code below is a simulation identical to the one just described, but with a scaring option included through `scaring = TRUE`.

@@

When scaring is introduced to an otherwise identical simulation (compare Figure 3 to Figure 1), the goose population increases as before, but it achieves the managers target population size and stabilises 2-3 time steps earlier. The reason for this earlier stabilisation is due to the change in farmer's actions as a consequence of the introduced scaring policy. At the start of the simulation, managers adjust policy by quickly increasing the cost of culling and decreasing the cost of scaring. In response, farmers turn to scaring rather than culling to remove geese from their land cells (Figure 3). This is in contrast to the simulation in which scaring was not an option, and farmers simply culled as much as possible despite the high costs (Figure 1). After the population has risen to slightly above the manager's target, the cost of culling again decreases, with the manager balancing the incentivisation of culling and scaring. The consequence of scaring as an available policy also reveals some potentially unexpected outcomes, such as increased variance in among-farmer agricultural production, which arises as geese are scared from one area of the landscape to another.

@@

Unlike the case in which culling was the only policy option (compare Figure 4 to Figure 2), the effort that farmers expended on a given action did not typically rise above the manager's permissiveness of the action. Hence, under the conditions of this model, the inclusion of scaring as a policy option has reduced conservation conflict in the social-ecological system. We again emphasise that the simulations presented here only serve as an example of how GMSE could be used as a tool for simulating social-ecological systems and understanding the potential for conflict; it is not intended to inform policy in Taiga Bean Geese or any other specific system.

@@

The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide set of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models. Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different policy and action possibilities. [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse). Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.

>>>>

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess conflict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer to simulate by [looping time steps](https://bradduthie.github.io/gmse_SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values. Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@Johnson2018 recently estimated key demographic parameters for the Taiga Bean Geese for the Central Management Unit that includes geese that breed in "Northern most Sweden, Northern Norway, Northern and Central Finland, and adjacent North-wester parts of Russia, wintering mostly in Southern Sweden and South-east Denmark" [@AEWA2016]. They estimated goose survival under ideal conditions to be ca 0.878; this can be interpreted in our model by setting mortality to `remove_pr = 1 - 0.878`. Similarly, @Johnson2018 estimated mean reproductive rate and carrying capacity to be 0.55 and 93870, respectively, so we set `lambda = 0.275` (for simplicity, we simply use half the mean reproductive rate; GMSE does not currently distinguish female and male individuals) and `res_death_K = 93870`. The global abundance of Taiga Bean Geese in 2009 was ca 63000 [@Fox2010], and ca 35000 in the Central Management Unit [@AEWA2016], which we can take as a starting abundance for our simulations (`RESOURCE_ini = 35000`). And the International Single Species Action Plan has a target population size of ca 70000 in the Central Management Unit [@AEWA2016], which we can use as a management target (`manage_target = 70000`). We simulate social-ecological dynamics over 30 time steps, which could be interpreted as years.

The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above. We also set `manager_budget = 100000` and `user_budget = 100000`. Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`). Landscape size is set to default 100 by 100 cells, so each farmer effectively owns 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the advanced GMSE options using `gmse_apply`). Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`. We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`). We further assume that the population is very well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe one landscape cell in every direction (`agent_view = 1`). All other parameter values are set to GMSE defaults.

@@

Below, we first only allow culling as a policy option and plot the dynamics of the social-ecological system from a single simulation. Next, we run the same simulation but also allow scaring as a policy option; we then use the model to make inferences regarding how scaring as a management option might affect goose population dynamics, agricultural production, and conservation conflict in the system. We emphasise that the simulations below are intended only to demonstrate one use of GMSE on a species of conservation interest, not to make recommendations for management of Taiga Bean Geese.

@@

Figure 1 shows the dynamics of goose abundance and agricultural yield, along with how managers react to change in abundance and farmers react to manager policy. In the case of the simulation above, managers quickly set a policy of high cost for culling, which leads to a rise in the goose population and a decrease in crop yield for farmers. After roughly 20 time steps, the goose population rises above the manager target, at which point the manager becomes more permissive of culling and the cost of culling for users therefore declines. In response, users begin to cull geese on their land, and the goose population begins to stabilise around the target of 7000 total geese. We can investigate the conflict between management policy and farmers more directly using the `plot_gmse_effort` function.

@@

Black lines indicate how permissive a manager is toward a particular action on a scale of 0 to 100, while coloured lines indicate how much effort farmers expend on a given action. When black lines are far below coloured lines, we can (cautiously) interpret this as a conflict between management of the goose population and farmer's interest in agricultural production. These time periods represent instances in which the manager is not permissive of a particular action (in this case culling), but farmers continue to expend effort to do the action anyway. In the case of the above simulation of potential conservation conflict between farmers and goose conservation, conflict is highest before time step 20, where the manager is not permissive of culling because the population is below the manager's target. Once the goose population has increased above the manager's target, conflict decreases because the desired culling is permitted by managers to keep the population at a target abundance. It is worth noting that, despite conflict as we define it decreasing, agricultural damage is still relatively high after the target goose population size is achieved (Figure 1). Hence, on a broader scale, conflict might persist around the appropriate target population size rather than what actions are permitted for farmers; currently, this potential aspect of conflict is not modelled, but future versions of GMSE may attempt to incorporate such additional complexity in conflict scenarios.

We can model the consequences for goose population dynamics, agricultural production, and conservation conflict when scaring is a policy option available to the manager. The code below runs a simulation identical to the one just discussed, but with a scaring option included using the argument `scaring = TRUE`.

@@

When scaring is introduced to an otherwise identical simulation (compare Figure 3 to Figure 1), the goose population increases as before, but it achieves the manager's target population size and stabilises 2-3 time steps earlier. The reason for this earlier stabilisation is due to the change in farmer's actions as a consequence of the introduced scaring policy. At the start of the simulation, managers adjust policy by quickly increasing the cost of culling and decreasing the cost of scaring. In response, farmers turn to scaring rather than culling to remove geese from their land cells (Figure 3). This is in contrast to the simulation in which scaring was not an option, and farmers simply culled as much as possible despite the high costs (Figure 1). After the population has risen to slightly above the manager's target, the cost of culling again decreases, with the manager balancing the incentivisation of culling and scaring. The consequence of scaring as an available policy also reveals some potentially unexpected outcomes, such as increased variance in among-farmer agricultural production, which arises as geese are scared from one area of the landscape to another.

@@

Unlike the case in which culling was the only policy option (compare Figure 4 to Figure 2), the effort that farmers expended on a given action did not rise as highly above the manager's permissiveness of the action. Hence, under the conditions of this model, the inclusion of scaring as a policy option has reduced conservation conflict in the social-ecological system. We again emphasise that the simulations presented here only serve as an example for how GMSE could be used as a tool for simulating social-ecological systems and understanding the potential for conflict; it is not intended to inform policy in Taiga Bean Geese or any other specific system.

@@

The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide a suite of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models. Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different manager policy and user action possibilities. [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse). Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems.
